---
title: "Red Wine Quality: Exploratory Data Analysis"
author: "by Filipe Tatarli"
#date: "21/02/2019"
output: html_document
---

* * *

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

  
  All the packages used in the analysis are listed below:
```{r Libraries, include=TRUE}
library(ggplot2)
library(gridExtra)
library(ggcorrplot)
```

```{r Data Preparation, include=FALSE}
wine.red <- read.csv('/Users/filipetatarli/Documents/R_Projects/data/wineQualityReds.csv')
```

# Univariate Plots Section  
  
  Our dataset consists of 13 variables, with 1.599 observations.  

```{r Data structure analysis, echo=FALSE}
str(wine.red)
```

```{r Data summary, echo=FALSE}
summary(wine.red)
```

```{r Create Quality Text, echo=FALSE}
wine.red$quality.cat <- NA
wine.red$quality.cat <- ifelse(wine.red$quality >= 7, 'good', NA)
wine.red$quality.cat <- ifelse(wine.red$quality == 5, 'medium', wine.red$quality.cat)
wine.red$quality.cat <- ifelse(wine.red$quality == 6, 'medium', wine.red$quality.cat)
wine.red$quality.cat <- ifelse(wine.red$quality <= 4, 'bad', wine.red$quality.cat)
wine.red$quality.cat <- factor(wine.red$quality.cat, levels = c("bad", "medium", "good"))
```

# Univariate Analysis  
  
  First of all I checked the distribution of the quality variable of the dataset where I noticed that most rated quality of red wines data are between 5 and 6. Besides that the assigned ratings range from 3 to 8.  
    
  
```{r Histogram of Wine Quality, echo=FALSE}
ggplot(aes(x = quality), data = wine.red) +
  geom_histogram(binwidth = 1, color = 'black', fill = '#F79420') +
  scale_x_continuous(breaks = seq(1,8,1)) +
  xlab('Quality') + 
  ylab('Count') +
  ggtitle('Histogram of Wine Quality')
  
table(wine.red$quality)
```
  
  Now I want to explore other attributes and understand if the variables appear in an normal distribution.  
    
    
```{r Univariate Analysis Graphs, echo=FALSE, warning=FALSE, message=FALSE}
p1 <- ggplot(aes(x = fixed.acidity), data = wine.red) +
  geom_histogram() +
  scale_x_continuous(limits = c(4, 16)) +
  ggtitle('Histogram of Fixed Acidity')

p1_log <- ggplot(aes(x = log(fixed.acidity)), data = wine.red) +
  geom_histogram() +
  ggtitle('Histogram of Fixed Acidity - Log')

p2 <- ggplot(aes(x = volatile.acidity), data = wine.red) +
  geom_histogram() +
  scale_x_continuous(limits = c(0, 2)) +
  ggtitle('Histogram of Volatile Acidity')

p2_log <- ggplot(aes(x = log(volatile.acidity)), data = wine.red) +
  geom_histogram() +
  ggtitle('Histogram of Volatile Acidity - Log')

p3 <- ggplot(aes(x = citric.acid), data = wine.red) +
  geom_histogram() +
  scale_x_continuous(limits = c(0, 1)) +
  ggtitle('Histogram of Citric Acid')

p3_log <- ggplot(aes(x = log(citric.acid)), data = wine.red) +
  geom_histogram() +
  ggtitle('Histogram of Citric Acid - Log')

p4 <- ggplot(aes(x = residual.sugar), data = wine.red) +
  geom_histogram() +
  scale_x_continuous(limits = c(0, 16)) +
  ggtitle('Histogram of Residual Sugar')

p4_log <- ggplot(aes(x = log(residual.sugar)), data = wine.red) +
  geom_histogram() +
  ggtitle('Histogram of Residual Sugar - Log')

p5 <- ggplot(aes(x = chlorides), data = wine.red) +
  geom_histogram() +
  scale_x_continuous(limits = c(0, 1)) +
  ggtitle('Histogram of Chlorides')

p5_log <- ggplot(aes(x = log(chlorides)), data = wine.red) +
  geom_histogram() +
  ggtitle('Histogram of Chlorides - Log')

p6 <- ggplot(aes(x = free.sulfur.dioxide), data = wine.red) +
  geom_histogram() +
  scale_x_continuous(limits = c(1, 72)) +
  ggtitle('Histogram of Free Sulfur Dioxide')

p6_log <- ggplot(aes(x = log(free.sulfur.dioxide)), data = wine.red) +
  geom_histogram() +
  ggtitle('Histogram of Free Sulfur Dioxide - Log')

p7 <- ggplot(aes(x = total.sulfur.dioxide), data = wine.red) +
  geom_histogram() +
  scale_x_continuous(limits = c(6, 289)) +
  ggtitle('Histogram of Total Sulfur Dioxide')

p7_log <- ggplot(aes(x = log(total.sulfur.dioxide)), data = wine.red) +
  geom_histogram() +
  ggtitle('Histogram of Total Sulfur Dioxide - Log')

p8 <- ggplot(aes(x = density), data = wine.red) +
  scale_x_continuous(limits = c(0, 1)) +
  geom_histogram() +
  ggtitle('Histogram of Density')

p8_log <- ggplot(aes(x = log(density)), data = wine.red) +
  geom_histogram() +
  ggtitle('Histogram of Density - Log')

p9 <- ggplot(aes(x = pH), data = wine.red) +
  scale_x_continuous(limits = c(2, 4)) +
  geom_histogram() +
  ggtitle('Histogram of pH')

p9_log <- ggplot(aes(x = log(pH)), data = wine.red) +
  geom_histogram() +
  ggtitle('Histogram of pH - Log')

p10 <- ggplot(aes(x = sulphates), data = wine.red) +
  scale_x_continuous(limits = c(0, 2)) +
  geom_histogram() +
  ggtitle('Histogram of Sulphates')

p10_log <- ggplot(aes(x = log(sulphates)), data = wine.red) +
  geom_histogram() +
  ggtitle('Histogram of Sulphates - Log')

p11 <- ggplot(aes(x = alcohol), data = wine.red) +
  scale_x_continuous(limits = c(8, 15)) +
  geom_histogram() +
  ggtitle('Histogram of Alcohol')

p11_log <- ggplot(aes(x = log(alcohol)), data = wine.red) +
  geom_histogram() +
  ggtitle('Histogram of Alcohol - Log')

grid.arrange(p1,  p1_log, ncol = 2)
grid.arrange(p2,  p2_log, ncol = 2)
grid.arrange(p3,  p3_log, ncol = 2)
grid.arrange(p4,  p4_log, ncol = 2)
grid.arrange(p5,  p5_log, ncol = 2)
grid.arrange(p6,  p6_log, ncol = 2)
grid.arrange(p7,  p7_log, ncol = 2)
  
```
  
#### What is the structure of your dataset?
  
  The dataset had __1599 observations__ with __13 features__ (fixed.acidity, volatile.acidity, citric.acid, residual.sugar, chlorides, free.sulfur.dioxide, total.sulfur.dioxide, density, pH, sulphates, alcohol and quality).
  
  
#### What is/are the main feature(s) of interest in your dataset?  
  
  The main feature of interest is the quality of the wine.  
  
  
#### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
  
  In univariate analysis, it is hard to find an variable that are related to quality.

#### Did you create any new variables from existing variables in the dataset?
  
  I created a variable for rating the wine quality, where:
  
* bad (<=4)
* medium (5 or 6)
* good ( >= 7)
  
  
#### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?  
  
  Most of the features had a normal distribution like density and pH with few outliers.  
Some features such as residual sugar and chlorides showed long tails what indicates the presence of outliers. 

# Bivariate Plots Section

# Bivariate Analysis

```{r Correlation, echo=FALSE}
corr <- round(cor(wine.red[, 2:13]),3)

ggcorrplot(corr, hc.order = TRUE, outline.col = "white", type = "lower", lab = TRUE, tl.cex = 7, digits = 1, tl.srt = 90)
```

#### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?  
  
  The variables alcohol and volatile acidity, have strong correlation with the quality. ( ___0.476___ and ___-0.391___)

#### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

  The variables fixed acidity and pH had a stronger negative correlation than others.

#### What was the strongest relationship you found?
  
  The strongest relationship was found betwenn fixed acidity and citric acid (r = ___0.672___).

# Multivariate Plots Section

# Multivariate Analysis
  
  Now we discovered that alchohol and quality has a considerable correlation (r = 0.476). We will investiguet other variables that can impact the quality. 

```{r Multivariate Plots v1, echo=FALSE}
ggplot(wine.red, aes(x = alcohol, y = volatile.acidity, color = quality)) +
  geom_jitter() +
  scale_color_gradient(high = 'green', low = 'red')

ggplot(wine.red, aes(x = alcohol, y = residual.sugar, color = quality)) +
  geom_jitter() +
  scale_color_gradient(high = 'green', low = 'red')

ggplot(wine.red, aes(x = alcohol, y = density, color = quality)) +
  geom_jitter() +
  scale_color_gradient(high = 'green', low = 'red')
```

#### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?
  
  I didn't find a strong relationship between the variables. I only observated that when the alcohol increases and the volitile acidity decreases we note that the wine quality increases.
  
#### Were there any interesting or surprising interactions between features?

  No

* * * 

# Final Plots and Summary

### Plot One

```{r Plots One, echo = FALSE}
ggplot(data = wine.red
       ,aes(x = quality.cat, y = alcohol, fill = quality.cat)) +
  geom_boxplot(show.legend = FALSE) +
  stat_summary(fun.y = mean, geom = 'point', shape = 4, show.legend = FALSE) +
  labs(title = "% ALCOHOL BY WINE QUALITY"
       ,x = "Quality"
       ,y = "% Alcohol")

by(wine.red$alcohol, wine.red$quality.cat, summary)
```

### Description One
  
  The boxplot shows how quality is affected by alcohol percentage. When higher the median better the quality. Besides that the medium quality has a lot of outliers. 

### Plot Two

```{r Plots Two, echo=FALSE}
ggplot(data = wine.red
       , aes(x = alcohol, y = quality)) +
  geom_jitter(alpha = 0.3, height = 0.5, width = 0.025) +
  geom_smooth(method = "lm") +
  scale_y_continuous(limits = c(4,8)) +
  ggtitle("Quality vs Alcohol") +
  xlab("Alcohol(%)") +
  ylab("Quality")
```


### Description Two
  
  In this plot we confirmed that when alcohol increases the quality of the wine increases too.

### Plot Three

```{r Plots Three, echo=FALSE}
ggplot(data = wine.red, aes(x = alcohol, y = density, color = factor(quality))) +
  geom_jitter() + # alpha
  ggtitle("Quality by Volitile Acidity and Alcohol") +
  xlab("Alcohol (%)") +
  ylab("Volitile Acidity (g/L)")

```


### Description Three  
  
  This graph shows the few values of better quality (8). Altought revels that que quality improves as the alcohol increases and the volitile decreases (red points).
  
* * *

# Reflection  
  
  In this data, my main struggle was to undestand the correlation between variables. I know nothing about wines. So I decide to understand each variable independent and then start building random graphs to understand the dataset.  
  I explored the interaction of every attribute with wine quality. I plot a lot of boxplots to detect the trends. 
  The plots and graphs make it easier to see where to explore more.  
  For future work, I intend on implement a predictive model with some statistical techniques like regression.

